Impact of Exam Format on Student Learning Outcomes

### 2019-05-13 CUT/paper.tex
Paper revisions

\usepackage{hyperref}
\usepackage{apacite}
\usepackage{setspace}
\usepackage[margin=1in]{geometry}
\doublespace
\title{Impact of Assessment Format on Student Learning Outcomes}
\author{W. Justin Toth\\Combinatorics and Optimization\\20411831}
\begin{document}
\maketitle
\begin{abstract}
We study of question of how assessment formats impact student study behaviours, learning outcomes, and perception of the courses they take. We particularly focus on Mathematics and Science courses, but the findings in this survey are general enough to be widely applicable. We study Multiple Choice versus Constructed Response, how much Reference Material to allow in the exam room, Oral versus Written Exams, and the Two Stage method for injecting collaborative learning into exams.
\end{abstract}
\newpage
\tableofcontents
\newpage
\section{Introduction}

@@

A well-known truism in human social interaction is that how you say something often has as much, if not more, importance that the specific content of your speech. Yet when instructors are designing evalutions for their courses, it is all too common that the focus is squarely on what questions to ask to cover the course content, rather than how the problems should be presented to the students. In this survey we seek to synthesize the academic literature on the ``how'', rather than ``what'', of assessment methods and provide guidance for instructors on the impact that exam format can have on student learning outcomes and their ability to measure the attainment thereof.

@@

In Section \ref{sec:written} we study the particulars of written finals. For written finals we will concentrate on two areas of focus in the literature. The first is on the use of multiple choice versus constructed response questions, and the second is on allowing students to use reference material such as the course textbook or handwritten notes during the exam. In Section \ref{sec:communication} we study the use of oral communication during exams. This breaks down in to two subsections. In Subsection \ref{subsec:oral} we survey the literature on implementing interview-style oral exams  between student and instructor in the classroom, and in Subsection \ref{subsec:collab} we discuss student-to-student collaboration in exams with a particular focus on the innovative two-stage exam format. We conclude by with Section \ref{sec:recommendations} where present recommendations for instructors on choosing exam formats for their courses based on the previously surveyed literature.

@@

Despite the advantages above, Multiple Choice questions have many disadvantages over Constructed Response questions. It is difficult to assess critical thinking skills with Multiple Choice questions~\cite{martinez1999cognition}, and it is essentially impossible to assess creativity with Multiple Choice questions. Students seem to be aware of this on some level, expressing a perception that Multiple Choice exams assess a more shallow, knowledge-level understanding of the course material~\cite{scouller1998influence}. In line with this viewpoint, students predominantly favor shallow-learning strategies like memorization for exams consisting primarily of Multiple Choice problems. In contrast, students perceive Construhcted Response exams as testing deeper understanding, and use deeper learning strategies to study for Constructed Response exams~\cite{scouller1998influence}

@@

The author defined critical-thinking skills as those skills associated with the middle levels of Bloom's taxonomy: application, analysis, evaluation, and synthesis. The experiment was performed using two sections of introductory biology given by the same instructor in the same semester to large classes of about $200$ students. The sections were entirely indentical in course material (lectures, assignments, readings) with the exception of the final exam. One section received a Multiple Choice only final, and the other received a Multiple Choice plus short answer Constructed Response final. Through a series of four surveys, Stanger-Hall evaluated student approaches to studying and perception of the importance of critical-thinking skills in introductory biology. The Constructed Response group reported significantly higher cognitively active learning behaviours during weeks outside of the exam study period, and this was reflected in their significantly higher test scores, in particular on questions associated with critical-thinking levels of Bloom's taxonomy~\cite{stanger2012multiple}.

@@

Another fundamental decision an instructor needs to make when designing an exam is whether or not they will allow students to make use of reference material during the exam. Perhaps the most common decision taken is to not allow reference material of any sort during an exam, the so-called Closed Book exam. On the other end of the reference material permissiveness spectrum is the Open Book exam. In an Open Book exam, students are permitted to use the primary reference material for the course, usually the recommended course textbook but alternatively all course notes may be permitted. For the purpose of our discussion an Open Book exam will be defined by the unmitigated allowance of a reference text where students can expect to find the answer to any reasonable basic fact about the subject they may need for an exam. Another option exists in the middle between Closed and Open Book exams. This option is what we will call Cheatsheet exams. In a Cheatsheet exam students are allowed to prepare a restricted set of notes to bring to the exam. Usually these notes are allowed to have any content the student may want on them, but with space and formatting restrictions which allow the notes they carry in to have significantly more compact information than an Open Book exam. Comparing the efficacy of Open versus Closed Book exams in the literature has led to mixed results.

@@

The anxiety Capalid noted is a common problem. Boedigheimer also observed that phenomenon. Another issue with Oral Exams is the instructor workload. In a class of $35$ to $40$ students, Oral Exams can easily span a couple of full days. Nevermind a first year course with $200$ plus students. This difficulty can spill over into missed class time, with the instructor often needing to bock off classes during the semester just for Oral Exams. 

@@

Oral Exams provide an intersting opportunity for student collaboration in the exam environment. \cite{odafe2006oral} gave an Oral Exam to students in a first Algebra course where students were tasked with solving a math problem in small groups. Odafe noted unique advantages and disadvantages to the group collaboration element. The obvious advantage is that students gain practice working on solutions as part of a team. This skill would be critical if they hoped to pursue mathematics as a career in the future, and is one that is not trained nearly enough in most students mathematical education. Teams that worked well together would have each student in the group contributing actively, and they were able to build off each others ideas learning and solving more in the exam than any one student would normally be able to. The disadvantages come with groups that did not work well together. If one member of a group was unable or unwilling to contribute, the students in that group were severely hampered by their teammate. Groups which poor communication or teamwork reported high dissatisfaction with the Oral Exam component of the course in the course evaluation surveys. Most students did not have that experience though, and overall student evaluations of the Oral Exam format were highly positive.

@@

We saw that while Multiple Choice exams are imminently pragmatic, they encourage shallow learning of course material, and are detrimental to long term retention when compared to Constructed Response exam problems. We recommend that instructors relegate Multiple Choice to assessing the lowest levels of Bloom's taxonomy, and even then to only use them in very large classes where the practical benefits outweigh the potential drawbacks.

>>>>

\usepackage{hyperref}
\usepackage{apacite}
\usepackage{setspace}
\usepackage[margin=1in]{geometry}
\doublespace
\title{Impact of Exam Format on Student Learning Outcomes}
\author{W. Justin Toth\\Combinatorics and Optimization\\20411831}
\begin{document}
\maketitle
\begin{abstract}
We study of question of how exam formats impact student study behaviours, learning outcomes, and perception of the courses they take. We particularly focus on Mathematics and Science courses, but the findings in this survey are general enough to be widely applicable. We study multiple choice versus constructed response, how much reference material to allow in the exam room, oral versus written exams, and the Two Stage method for injecting collaborative learning into exams.
\end{abstract}
\newpage
\tableofcontents
\newpage
\section{Introduction}

@@

A well-known truism in human social interaction is that how you say something often has as much, if not more, importance that the specific content of your speech. Yet when instructors are designing exams for their courses, it is all too common that the focus is squarely on what questions to ask to cover the course content, rather than how the problems should be presented to the students. In this survey we seek to synthesize the academic literature on the ``how'', rather than ``what'', of assessment methods used for exams and provide guidance for instructors on the impact that exam format can have on student learning outcomes and their ability to measure the attainment thereof.

@@

In Section \ref{sec:written} we study the particulars of written finals. We concentrate on exams in the context of STEM courses, but many of the findings are broadly applicable. For written finals we will concentrate on two areas of focus in the literature. The first is on the use of multiple choice versus constructed response questions, and the second is on allowing students to use reference material such as the course textbook or handwritten notes during the exam. In Section \ref{sec:communication} we study the use of oral communication during exams. This breaks down in to two subsections. In Subsection \ref{subsec:oral} we survey the literature on implementing interview-style oral exams  between student and instructor in the classroom, and in Subsection \ref{subsec:collab} we discuss student-to-student collaboration in exams with a particular focus on the innovative two-stage exam format. We conclude by with Section \ref{sec:recommendations} where present recommendations for instructors on choosing exam formats for their courses based on the previously surveyed literature.

@@

Despite the advantages above, Multiple Choice questions have many disadvantages over Constructed Response questions. It is difficult to assess critical thinking skills with Multiple Choice questions~\cite{martinez1999cognition}, and it is essentially impossible to assess creativity with Multiple Choice questions. Students seem to be aware of this on some level, expressing a perception that Multiple Choice exams assess a more shallow, knowledge-level understanding of the course material~\cite{scouller1998influence}. In line with this viewpoint, students predominantly favor shallow-learning strategies like memorization for exams consisting primarily of Multiple Choice problems. In contrast, students perceive Constructed Response exams as testing deeper understanding, and use deeper learning strategies to study for Constructed Response exams~\cite{scouller1998influence}

@@

The author defined critical-thinking skills as those skills associated with the middle and high levels of Bloom's taxonomy: application, analysis, evaluation, and synthesis. The experiment was performed using two sections of introductory biology given by the same instructor in the same semester to large classes of about $200$ students. The sections were entirely indentical in course material (lectures, assignments, readings) with the exception of the final exam. One section received a Multiple Choice only final, and the other received a Multiple Choice plus short answer Constructed Response final. Through a series of four surveys, Stanger-Hall evaluated student approaches to studying and perception of the importance of critical-thinking skills in introductory biology. The Constructed Response group reported significantly higher cognitively active learning behaviours during weeks outside of the exam study period, and this was reflected in their significantly higher test scores, in particular on questions associated with critical-thinking levels of Bloom's taxonomy~\cite{stanger2012multiple}.

@@

Another fundamental decision an instructor needs to make when designing an exam is whether or not they will allow students to make use of reference material during the exam. Perhaps the most common decision taken is to not allow reference material of any sort during an exam, the so-called Closed Book exam. On the other end of the reference material permissiveness spectrum is the Open Book exam. In an Open Book exam, students are permitted to use the primary reference material for the course, usually the recommended course textbook but alternatively all course notes may be permitted. For the purpose of our discussion an Open Book exam will be defined by the unmitigated allowance of a reference text where students can expect to find the answer to any reasonable basic fact about the subject they may need for an exam. Another option exists in the middle between Closed and Open Book exams. This option is what we will call Cheatsheet exams. In a Cheatsheet exam students are allowed to prepare a restricted set of notes to bring to the exam. Usually these notes are allowed to have any content the student may want on them, but with space and formatting restrictions which allow the notes they carry in to have significantly more compact information than an Open Book exam. Comparing the overall efficacy, in terms of encouraging postive study habits and long term knowledge retention, of Open versus Closed Book exams in the literature has led to mixed results.

@@

The anxiety Capaldi noted is a common problem. Boedigheimer also observed that same phenomenon. Another issue with Oral Exams is the instructor workload. In a class of $35$ to $40$ students, Oral Exams can easily span a couple of full days. Nevermind a first year course with $200$ plus students where the workload can become entirely infeasible without considerable instructional support. This difficulty can spill over into missed class time, with the instructor often needing to block off classes during the semester just for Oral Exams. 

@@

Oral Exams provide an intersting opportunity for student collaboration in the exam environment. \cite{odafe2006oral} gave an Oral Exam to students in a first Algebra course where students were tasked with solving a math problem in small groups. Odafe noted unique advantages and disadvantages to the group collaboration element. The obvious advantage is that students gain practice working on solutions as part of a team. This skill would be critical if they hoped to pursue mathematics as a career in the future, and is one that is not trained nearly enough in most students mathematical education. Teams that worked well together would have each student in the group contributing actively, and they were able to build off each others ideas learning and solving more in the exam than any one student would normally be able to. The disadvantages come with groups that did not work well together. If one member of a group was unable or unwilling to contribute, the students in that group were severely hampered by their teammate. Groups with poor communication or teamwork reported high dissatisfaction with the Oral Exam component of the course in the course evaluation surveys. Most students did not have that experience though, and overall student evaluations of the Oral Exam format were highly positive.

@@

We saw that while Multiple Choice exams are imminently pragmatic, they encourage shallow learning of course material, and are detrimental to long term retention when compared to Constructed Response exam problems. We recommend that instructors relegate Multiple Choice to assessing the lowest levels of Bloom's taxonomy, and even then to only use them in very large classes where the practical benefits outweigh the potential drawbacks. If an instructor is in the situation where Multiple Choice exams are being used, and they still desire to test higher-order learning objectives there are some best-practices avaiable in the literature. See for example in the context of teaching Biology~\cite{crowe2008biology}, or for Economics~\cite{buckles2006using}.
